Clustering time series based on probability distributions across temporal granularities

### 2021-10-31 main.Rmd
methodology flowchart introduced

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data.The similarity measure is the most essential ingredient of time series clustering. First step is to decide what we mean by similar. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from different data generating processes. Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. Each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered, there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. As a result, we need to decide how to measure similarities between these collections of univariate probability distributions. There are multiple ways to measure similarities between time series based on probability distributions, depending on the objective of the problem. This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. It considers two methodologies for finding distances between time series, which are explained in the next segment.

\noindent The sum of the distances between two series $x$ and $y$ in terms of cyclic granularity $A$ is defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity $A$) or $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. After determining the distance between two series in terms of one granularity, we must combine them to produce a distance based on all significant granularities. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels,  $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is employed, which is also a distance metric since it is the sum of JS distances.

<!-- Once we have chosen relevant features to characterize the probability distributions, we have to decide how to combine them into a single number, which serves as a distance between two observations. -->

<!-- which are described below. -->

<!-- Consequently, the distance measure might also change depending on the similarity measure chosen. -->

<!-- The (dis) similarity measure in this paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities and considers two methodology for finding distances between observations, which are explained in the next segment. -->

It is worth noting that when studying these similarities, a variety of objectives may be pursued. One objective could be to group time series with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. There
are distance measures are used for shape-based clustering [Ding et al. 2008; Wang
et al. 2013] and many more but none of them look at the probability distributions while computing similarity. Moreover, most distance measures offer similar shape across just one dimension. For example, we often see "similar" daily energy profiles across hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar distributional shape across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:

@@

This distance matrix considers two objects to be similar if every category of an individual cyclic granularity or combination of categories for interacting cyclic granularities have similar distributions. In this study, the distribution for each category is characterized using deciles and the distances between distributions are computed by using the Jensen-Shannon distance, which is symmetric and hence could be used as a distance measure.

\noindent The total distance between two elements $x$ and $y$ is then defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity A) or  $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels, $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is used, which is also a distance metric being the sum of JS distances.

@@

Compute weighted pairwise distances ($wpd$) for all considered granularities for all objects. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting pattern is expected, whereas a lower value would indicate otherwise.

\noindent Distance between elements is then taken as the euclidean distances between them with the granularities being the variables and $wpd$ being the value under each variable. Since Euclidean distance is chosen, the observations with high values of features ($wpd$ values) will be clustered together. The same holds true for observations with low values of features. Thus this distance matrix would be useful to group customers that have similar significance of patterns across different granularities.

>>>>

The proposed methodology aim to leverage the intrinsic temporal data structure hidden in time series data. The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data.The similarity measure is the most essential ingredient of time series clustering. First step is to decide what we mean by similar. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand. In this work, instead of considering the probability distributions of the linear time series, we assume that the measured variables across different categories of any cyclic granularity are from different data generating processes. Hence, we want to be able to cluster distributions of the form $f_{i,A,B \dots, {N_C}}(v)$, where $A, B$ represent the cyclic granularities under consideration such that $A = \{a_j: j=1, 2, \dots J\}$,  $B = \{b_k: k  = 1, 2, \dots K\}$ and so on. We consider individual category of a cyclic granularity ($A$) or combination of categories for interaction of cyclic granularities (for e.g. $A*B$) to have a distribution. For example, let us consider we have two cyclic  granularities of interest, $A = {0, 1, 2, \dots, 23}$ representing hour-of-day and  $B = \{Mon, Tue, Wed, \dots, Sun\}$ representing day-of-week. Each customer $i$ consist of a collection of probability distributions. In case individual granularities ($A$ or $B$) are considered, there are  $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. As a result, we need to decide how to measure similarities between these collections of univariate probability distributions. There are multiple ways to measure similarities between time series based on probability distributions, depending on the objective of the problem. This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. It considers a methodology with two appraoches for  finding distances between time series. Both of these appraoches may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. The obtained distances could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. The methodology is explained in the Figure \ref{fig:flowchart} and each element of the pipeline is discussed.

```{r flowchart, fig.cap = "Flow chart illustrating the pipeline for methodology", out.width="100%"}
knitr::include_graphics("figs/flowchart.png")

<!-- It is worth noting that when studying these similarities, a variety of objectives may be pursued. One objective could be to group time series with similar shapes over all relevant cyclic granularities. In this scenario, the variation in customers within each group is in magnitude rather than shape, while the variation between groups is only in shape. There -->
<!-- are distance measures are used for shape-based clustering [Ding et al. 2008; Wang -->
<!-- et al. 2013] and many more but none of them look at the probability distributions while computing similarity. Moreover, most distance measures offer similar shape across just one dimension. For example, we often see "similar" daily energy profiles across hours of the day, but we suggest a broader approach to the problem, aiming to group consumers with similar distributional shape across all significant cyclic granularities. Another purpose of clustering could be to group customers that have similar differences in patterns across all major cyclic granularities, capturing similar jumps across categories regardless of the overall shape. For example, in the first goal, similar shapes across hours of the day will be grouped together, resulting in customers with similar behaviour across all hours of the day, whereas in the second goal, any similar big-enough jumps across hours of the day will be clubbed together, regardless of which hour of the day it is. Both of these objectives may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. Depending on the goal of clustering, the distance metric for defining similarity would be different. These distance metrics could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this. This section presents the work flow of the methodology:  -->

@@

\noindent The sum of the distances between two series $x$ and $y$ in terms of cyclic granularity $A$ is defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity $A$) or $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. After determining the distance between two series in terms of one granularity, we must combine them to produce a distance based on all significant granularities. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels,  $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is employed, which is also a distance metric since it is the sum of JS distances.  In this approach, the variation in time series within each group is in magnitude rather than distributional pattern, while the variation between groups is only in distributional pattern across categories.

@@
